RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-11-21 manuscripts/manuscript.Rmd
Emphasize semantics in the intro
and lots of other revisions trying to be more clear about the advantages of NeXML over NEXUS.

- Introduction
  - Exciting times in phylogenetics!  
  - Challenges with existing formats 
  - Overview of NeXML
  - Map

@@

Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible format to exchange and archive comparative methods data.
A common file format for sharing phylogenetic tree data has long played
an essential role in the advancement of phylogenetics research.
The rapid increase in phylogenetics data and research methods that make
use of such data has introduced many new challenges and opportunities 
to sharing this information both between software environments and 
between research groups [@Drew_2013; @Stoltzfus_2012].

The ever expanding needs of the phylogenetics community are rapidly
out-pacing the capacity of current standards, which is resulting in a
radiation of different data representations and a loss of interoperability
at the very time when the growth of available data and methods has made
that interoperability most valuable.

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led developers to
repeatedly add small modifications to the standard to meet the needs
of their own software, which results in mutually incompatible dialects
that largely defeat the advantages of a shared  file format.
NeXML was developed to addresses each of these problems [@Vos_2012]  
by drawing on robust web standards in data representation.
The design and expressiveness of NeXML makes it much easier to meet
the rising bar for best practices in data sharing in phylogenetic 
research [@Cranston_2014].

The R language [@R] has become an important environment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations [@taskview],
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `ape::phylo` format provided by the
ape package [@Paradis_2004].

Though certain functions allow R users to switch between these formats
within R, these formats frequently contain more information than their
NEXUS file counterparts,
<!-- such as what? we haven't talked about metadata yet. HL -->
and users lack a suitable alternative for
archiving and sharing phylogenetic data outside of the R ecosystem. 
Consequently, information added to phylogenies by these R programs cannot 
easily be shared, archived, and made available to other software;
<!-- such as what? should walk through an example here. HL -->
but inherits the limitations of the NEXUS format.
<!-- how can information inherit the limitations of a serialization -->
<!-- format if it's added to data read from that format? HL -->
<!-- CB: Sorry, was refering to data _not_ read from NEXUS format,
but rather data created in an R format/data-structure.  Information
that is in that R data structure (e.g. a simmap) is lost when using
write.nexus(simmap) -->

`RNeXML` opens up a rich world of both currently available and rapidly
growing data sets in the NeXML format to the R user and the large ecosystem
of R packages currently available for handling phylogenetic data. 
R packages and R users frequently add rich information to the phylogenies
they work with, but need a platform independent data exchange format to
share and archive this. We illustrate how NeXML can
be used to distribute both phylogenies and character trait data used in 
many popular phylogenetic comparative methods tools in R.

`RNeXML` also provides convenient tools for adding metadata
to NeXML files, including the automated generation of metadata and examples
of how to extend the NeXML standard to represent new data features without
breaking validation or backwards compatibility of the file type.

<!-- FIXME something about how we will increasingly want to work with (semantic) metadata
and how RNeXML provides a solid platform from which to do so.
-->
<!-- I think this is in fact more important than the paragraph above. I.e., -->
<!-- whether exchanging basic comparative data needs another format is -->
<!-- debatable, but that metadata that can hardly be expressed in -->
<!-- current standards does is a clear point. HL -->

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in data structures
provided by existing R packages
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

>>>>

- Introduction
  - Exciting times in phylogenetics!  
  - Challenges with existing formats 
  - Overview of NeXML
  - Challenges with R
  - Map

@@

Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible, and semantically aware format to import, exchange and archive 
phylogenetic trees, traits, and metadata.
The rapid increase in phylogenetics data and research methods that make
use of such data has introduced many new challenges and opportunities 
for sharing, both between software environments and 
between research groups [@Drew_2013; @Stoltzfus_2012].
The ever expanding needs of the phylogenetics community are rapidly
out-pacing the capacity of current standards, which is resulting in a
radiation of different data representations and a loss of interoperability
at the very time when the growth of available data and methods has made
that interoperability most valuable.

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led many developers to
modify NEXUS in different ways for different needs, with the unfortunate
result that NEXUS files generated by one program can be incompatible
with another [@Vos_2012]. Without a formal grammar, software based on 
NEXUS files may also make inconsistent assumptions about tokens, quoting,
or element lengths.  @Vos_2012 estimates as many as 15% of the NEXUS
files in the CIPRES portal contain unrecoverable but hard to diagnose errors.

NeXML was developed to addresses each of these problems and also open doors
to new opportunities in handling phylogenetic data. First, NeXML is defined
by a precise grammar that can be programmatically **validated**. This helps
ensure that any NeXML file can always be read without errors by any 
NeXML parsing software. Second, NeXML is **extensible**: a user can define
representations of new information (as we will illustrate) without breaking
this compatibility. Third and most importantly, NeXML is **semantically rich**.
The use of formal semantics in NeXML provides a representation of data that
is not only machine-readable, but can also permits machine reasoning about
the data provided (see @Parr2011 for an overview on semantic reasoning in 
evolutionary informatics). Instead of asking the computer to
return "the data on line #93", we can expect the computer to reason about
taxonomic concepts (as we illustrate in the supplement).

The R language [@R] has become an important environment for phylogenetics
research, with over 30 packages available including several with over 
1000 citations [@taskview]. `RNeXML` makes the large and rapidly growing
archive of richly annotated, NeXML-based phylogenetics data available to
the R user and this powerful ecosystem of R tools for phylogenetics.
Further, R packages and users frequently add their own additional 
information to the phylogenies they work with: annotations of species,
stochastic character maps, trait values, model estimates and parameter 
values. By leveraging the semantic and extensible nature of NeXML, `RNeXML` 
provides R users a platform independent data exchange format to capture,
further annotate, exchange, and preserve this information. With `RNeXML`,
R users and developers can both better contribute to and take advantage 
of the rapidly growing infrastructure of evolutionary informatics [@Parr2011].  
By capturing information already available in the R environment, together
with the appropriate choice of defaults, `RNeXML` can facilitate meeting
the rising bar for best practices in data sharing in phylogenetic 
research [@Cranston_2014] while at the same reducing the burden
on the user.

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses semantic metadata to
extend the NeXML standard to describe stochastic character mapping: a kind of
data annotation that is currently available in data structures
provided by existing R packages
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.
